\pkg{NeuralNetTools}: Visualization and Analysis Tools for Neural Networks

### 2018-05-03 final_JSS_edits/paper.tex
working on final JSS edits

\documentclass[article,shortnames]{jss}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% declarations for jss.cls %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% almost as usual
\author{Marcus W. Beck\\US Environmental Protection Agency}
\title{\pkg{NeuralNetTools}: Visualization and Analysis Tools for Neural Networks}

%% for pretty printing and a nice hypersummary also set:
\Plainauthor{Marcus W. Beck} %% comma-separated
\Plaintitle{NeuralNetTools: Visualization and Analysis Tools for Neural Networks} %% without formatting
\Shorttitle{\pkg{NeuralNetTools}: Visualization and Analysis Tools for Neural Networks}

%% an abstract and keywords
\Abstract{
Supervised neural networks have been applied as a machine learning technique to identify and predict emergent patterns among multiple variables.  A common criticism of these methods is the inability to characterize relationships among variables from a fitted model. Although several techniques have been proposed to `illuminate the black box', they have not been made available in an open-source programming environment.  This article describes the \pkg{NeuralNetTools} package that can be used for the interpretation of supervised neural network models created in \proglang{R}.  Functions in the package can be used to visualize a model using a neural network interpretation diagram, evaluate variable importance by disaggregating the model weights, and perform a sensitivity analysis of the response variables to changes in the input variables.  Methods are provided for objects from many of the common neural network packages in \proglang{R}, including \pkg{caret}, \pkg{neuralnet}, \pkg{nnet}, and \pkg{RSNNS}.  The article provides a brief overview of the theoretical foundation of neural networks, a description of the package structure and functions, and an applied example to provide a context for model development with \pkg{NeuralNetTools}.  Overall, the package provides a toolset for neural networks that complements existing quantitative techniques for data-intensive exploration.      
}
\Keywords{neural networks, \code{plotnet}, sensitivity, variable importance, \proglang{R}}
\Plainkeywords{neural networks, plotnet, sensitivity, variable importance, R} %% without formatting
%% at least one keyword must be supplied

%% publication information
%% NOTE: Typically, this can be left commented and will be filled out by the technical editor
%% \Volume{50}
%% \Issue{9}
%% \Month{June}
%% \Year{2012}
%% \Submitdate{2012-06-04}
%% \Acceptdate{2012-06-04}

%% The address of (at least) one author should be given
%% in the following format:
\Address{
  Marcus W. Beck\\
  US Environmental Protection Agency\\
  National Health and Environmental Effects Research Laboratory\\
  Gulf Ecology Division, 1 Sabine Island Drive\\
  Gulf Breeze, Florida, 32561, USA\\
  Telephone: +1/850/934-2480\\
  E-mail: \email{beck.marcus@epa.gov}
}

% options and libs
% get online bib file

%% need no \usepackage{Sweave.sty}

@@

%% Note that you should use the \pkg{}, \proglang{} and \code{} commands.

@@

A common objective of data-intensive analysis is the synthesis of unstructured information to identify patterns or trends `born from the data' \citep{Bell09,Kelling09,Michener12}.  Analysis is primarily focused on data exploration and prediction as compared to hypothesis-testing using domain-specific methods for scientific exploration \citep{Kell03}.  Demand for quantitative toolsets to address challenges in data-rich environments has increased drastically with the advancement of techniques for rapid acquisition of data. Fields of research characterized by high-throughput data (e.g., bioinformatics, \citet{Saeys07}) have a strong foundation in computationally-intensive methods of analysis, whereas disciplines that have historically been limited by data quantity (e.g., field ecology, \citet{Swanson15}) have also realized the importance of quantitative toolsets given the development of novel techniques to acquire information.  Quantitative methods that facilitate inductive reasoning can serve a complementary role to conventional, hypothesis-driven approaches to scientific discovery \citep{Kell03}.

Statistical methods that have been used to support data exploration are numerous \citep{Jain00,Recknagel06,Zuur10}.  A common theme among data intensive methods is the use of machine-learning algorithms where the primary objective is to identify emergent patterns with minimal human intervention.  Neural networks, in particular, are designed to mimic the neuronal structure of the human brain by `learning' inherent data structures through adaptive algorithms \citep{Rumelhart86,Ripley96}.  Although the conceptual model was introduced several decades ago \citep{McCulloch43}, neural networks have had a central role in emerging fields focused on data exploration.  The most popular form of neural network is the feed-forward \ac{mlp} trained using the backpropagation algorithm \citep{Rumelhart86}.  This model is typically used to predict the response of one or more variables given one to many explanatory variables.  The hallmark feature of the \ac{mlp} is the characterization of relationships using an arbitrary number of parameters (i.e., the hidden layer) that are chosen through iterative training with the backpropagation algorithm.  Conceptually, the \ac{mlp} is a hyper-parameterized non-linear model that can fit a smooth function to any dataset with minimal residual error \citep{Hornik91}.

An arbitrarily large number of parameters to fit a neural network provides obvious predictive advantages, but complicates the extraction of model information.  Diagnostic information such as variable importance or model sensitivity are necessary aspects of exploratory data analysis that are not easily obtained from a neural network. As such, a common criticism is that neural networks are `black boxes' that offer minimal insight into relationships among variables \citep[e.g.,][]{Paruelo97}.  \citet{Olden02} provide a rebuttal to this concern by describing methods to extract information about variable relationships from neural networks.  Many of these methods were previously described but not commonly used.  For example, \citet{Olden02} describe  the \ac{nid} for plotting \citep{Ozesmi99}, the Garson algorithm for variable importance \citep{Garson91}, and the Profile method for sensitivity analysis \citep{Lek96}.  These quantitative tools `illuminate the black box' by disaggregating the network parameters to characterize relationships between variables that are described by the model.  Although \ac{mlp} neural networks were developed for prediction, methods described in \citet{Olden02} leverage these models to describe data signals.  Increasing the accesibility of these diagnostic tools will have value for exploratory data analysis and may also inform causal inference.

This article describes the \pkg{NeuralNetTools} package for \proglang{R} that was developed to better understand information obtained from the \ac{mlp} neural network.  Functions provided by the package are those described in \citet{Olden02} but have not been previously available in an open-source programming environment.  The reach of the package is extensive in that generic functions were developed for model objects from the most popular neural network packages available in \proglang{R}.  The objectives of this article are to 1) provide an overview of the statistical foundation of the \ac{mlp} network, 2) describe the theory and application of the main functions in the \pkg{NeuralNetTools} package, and 3) provide an applied example using neural networks and \pkg{NeuralNetTools} in data exploration.  The package is currently available on the \ac{CRAN}, whereas the development version is maintained as a GitHub repository.

\section[Theoretical foundation]{Theoretical foundation and existing \proglang{R} packages}

The typical \ac{mlp} network is composed of multiple layers that define the transfer of information between input and response layers.  Information travels in one direction where a set of values for variables in the input layer propagates through one or more hidden layers to the final layer of the response variables. Hidden layers between the input and response layers are key components of a neural network that mediate the transfer of information.  Just as the input and response layers are composed of variables or nodes, each hidden layer is composed of nodes with weighted connections that define the strength of information flow between layers.  Bias layers connected to hidden and response layers may also be used that are analagous to intercept terms in a standard regression model.

Training a neural network model requires identifying the optimal weights that define the connections between the model layers.  The optimal weights are those that minimize prediction error for a test dataset that is independent of the training dataset.  Training is commonly achieved using the backpropagation algorithm described in \citet{Rumelhart86}.  This algorithm identifies the optimal weighting scheme through an iterative process where weights are gradually changed through a forward- and backward-propagation process \citep{Rumelhart86,Lek00}.  The algorithm begins by assigning an arbitrary weighting scheme to the connections in the network, followed by estimating the output in the response variable through the forward-propagation of information through the network, and finally calculating the difference between the predicted and actual value of the response.  The weights are then changed through a backpropagation step that begins by changing weights in the output layer and then the remaining hidden layers.  The process is repeated until the chosen error function is minimized, as in standard model-fitting techniques for regression \citep{Cheng94}.  A fitted \ac{mlp} neural network can be represented as \citep{Bishop95,Venables02}:
\begin{equation}
y_k = f_o \left(\sum\limits_{h} w_{hk}f_h \left( \sum\limits_{i} w_{ih}x_i\right) \right)
\end{equation}
where the estimated value $k$ of the response variable $y$ is a sum of products between the respective weights $w$ for $i$ input variables $x$ and $h$ hidden nodes, mediated by the activation functions $f_h$ and $f_o$ for each hidden and output node.

Methods in \pkg{NeuralNetTools} were written for several \proglang{R} packages that can be used to create \ac{mlp} neural networks: \pkg{neuralnet} \citep{Fritsch12}, \pkg{nnet} \citep{Venables02}, and \pkg{RSNNS} \citep{Bergmeir12}. Limited methods were also developed for neural network objects created with the \code{train} function from the \pkg{caret} package \citep{Kuhn15}.  Additional \proglang{R} packages that can create \ac{mlp} neural networks include \pkg{AMORE} that implements the `TAO-robust backpropagation algorithm' for model fitting \citep{Castejon14}, \pkg{FCNN4R} as an \proglang{R} interface to the \pkg{FCNN} \proglang{C++} library \citep{Klima15}, \pkg{monmlp} for networks with partial monotonicity constraints \citep{Cannon15}, and \pkg{qrnn} for quantile regression neural networks \citep{Cannon11}.  At the time of writing, the \ac{CRAN} download logs \citep{Csardi15} showed that the \proglang{R} packages with methods in \pkg{NeuralNetTools} included 95\% of all downloads for the available \ac{mlp} packages, with \pkg{nnet} accounting for over 78\%.  As such, methods have not been included in \pkg{NeuralNetTools} for the remaining packages, although further development of \pkg{NeuralNetTools} could include additional methods based on popularity.  Methods for each function are currently available for  \code{mlp} (\pkg{RSNNS}), \code{nn} (\pkg{neuralnet}), \code{nnet} (\pkg{nnet}), and \code{train} (\pkg{caret}, only if the object also inherits from the \code{nnet} class).  Additional \code{default} or \code{numeric} methods are available for some of the generic functions.

@@

The stable release of \pkg{NeuralNetTools} can be installed from \ac{CRAN} and loaded as follows:

\begin{Schunk}
\begin{Sinput}
R> install.packages("NeuralNetTools")
R> library("NeuralNetTools")
\end{Sinput}
\end{Schunk}

\pkg{NeuralNetTools} includes four main functions that were developed following similar techniques in \citet{Olden02} and references therein.  Functions include \code{plotnet} to plot a neural network interpretation diagram, \code{garson} and \code{olden} functions to evaluate variable importance, and \code{lekprofile} for a sensitivity analysis of neural network response to input variables.  Most of the functions require the extraction of model weights in a common format for the neural network object classes in \proglang{R}.  The \code{neuralweights} function can be used to retrieve model weights for any of the model classes described above.  A two-element \code{list} is returned with the first element describing the structure of the network (number of nodes in the input, hidden, and output layers) and the second element as a named list of model weights.  The function is used internally within the main functions but may be useful for comparing networks of different classes.

A common approach for data pre-processing is to normalize the input variables and to standardize the response variables \citep{Lek00,Olden02}.  A sample dataset that follows this format is included with \pkg{NeuralNetTools}.  The \code{neuraldat} dataset is a simple \code{data.frame} with 2000 rows of observations and five columns for two response variables (\code{Y1} and \code{Y2}) and three input variables (\code{X1}, \code{X2}, and \code{X3}).  The input variables are random observations from a standard normal distribution and the response variables are linear combinations of the input variables with additional random components.  The response variables are also standardized from zero to one.  Additional datasets can be pre-processed to this common format using the \code{scale} function from \pkg{base} and the \code{rescale} function from \pkg{scales} for the input and response variables, respectively.  The examples below use three models created from the \code{neuraldat} dataset and include \code{mlp} (\pkg{RSNNS}), \code{nn} (\pkg{RSNNS}), and \code{nnet} (\pkg{nnet}) objects.

\begin{Schunk}
\begin{Sinput}
R> set.seed(123)
R> library("RSNNS")
R> x <- neuraldat[, c('X1', 'X2', 'X3')]
R> y <- neuraldat[, 'Y1']
R> mod1 <- mlp(x, y, size = 5)
R> library("neuralnet")
R> mod2 <- neuralnet(Y1 ~ X1 + X2 + X3, data = neuraldat, hidden = 5)
R> library("nnet")
R> mod3 <- nnet(Y1 ~ X1 + X2 + X3, data = neuraldat, size = 5)
\end{Sinput}
\end{Schunk}

@@

Tbe \code{plotnet} function in \pkg{NeuralNetTools} plots a \acl{nid} \citep[\acs{nid},][]{Ozesmi99} and includes several options to customize aesthetics. A \ac{nid} is a modification of the standard conceptual illustration of the \ac{mlp} network that changes the thickness and color of the weight connections based on magnitude and sign, respectively.  Positive weights between layers are shown as black lines and negative weights as grey lines. Line thickness is proportional to the absolute magnitude of each weight (\cref{fig:plotnet}).

\begin{figure}[!ht]
\subfloat[][]{
\includegraphics[width=7.5cm,page=1]{figs/plotnet.pdf}
\label{fig:plotnet1}
}
\subfloat[][]{
\includegraphics[width=7.5cm, page=2]{figs/plotnet.pdf}
\label{fig:plotnet2}
}
\caption{Examples from the \code{plotnet} function showing neural networks as a standard graphic (\ref{fig:plotnet1}) and using the \acl{nid} (\ref{fig:plotnet2}).  Labels outside of the nodes represent variable names and labels within the nodes indicate the layer and node (I: input, H: hidden, O: output, B: bias).}
\label{fig:plotnet}
\end{figure}

A primary and skip layer network can also be plotted for \code{nnet} models with a skip layer connection (\cref{fig:plotnet_skip}). Models with skip layers include additional connections from the input to output layers that bypass the hidden layer \citep{Ripley96}.  The default behavior of \code{plotnet} is to plot the primary network, whereas the skip layer can be viewed separately with \code{skip = TRUE}. If \code{nid = TRUE}, the line widths for both the primary and skip layer plots are relative to all weights. Plotting a network with only a skip layer (i.e., no hidden layer, \code{size = 0} in \code{nnet}) will include bias connections to the output layer, whereas these are included only in the primary plot if size is greater than zero.

\begin{figure}[!ht]
\subfloat[][]{
\includegraphics[width=7.5cm,page=1]{figs/plotnet_skip.pdf}
\label{fig:plotnet_skip1}
}
\subfloat[][]{
\includegraphics[width=7.5cm, page=2]{figs/plotnet_skip.pdf}
\label{fig:plotnet_skip2}
}
\caption{Examples from the \code{plotnet} function showing a neural network with a separate skip layer between the input and output layers.  The skip layer (\ref{fig:plotnet_skip1}) and primary neural network (\ref{fig:plotnet_skip2}) can be viewed separately with \code{plotnet} by using \code{skip = TRUE} or \code{skip = FALSE}.}
\label{fig:plotnet_skip}
\end{figure}

The \pkg{RSNNS} package provides algorithms to prune connections or nodes in a neural network \citep{Bergmeir12}.  This approach can remove connection weights between layers or input nodes that do not contribute to the predictive performance of the network.  In addition to visualizing connections in the network that are not important, connections that are pruned can be removed in successive model fitting.  This reduces the number of free parameters (weights) that are  estimated by the model optimization algorithm, increasing the likelihood of convergence to an estimable numeric solution for the remaining connection weights that minimizes prediction error \citep[i.e., model identifiability,][]{Ellenius00}. Algorithms in \pkg{RSNNS} for weight pruning include magnitude-based pruning, Optimal Brain Damage, and Optimal Brain Surgeon, whereas algorithms for node pruning include skeletonization and the non-contributing units method \citep{Zell98}.  The \code{plotnet} function can plot a pruned neural network, with options to omit or display the pruned connections (\cref{fig:plotprune}).

\begin{Schunk}
\begin{Sinput}
R> pruneFuncParams <- list(max_pr_error_increase = 10.0, pr_accepted_error = 1.0,
+   no_of_pr_retrain_cycles = 1000, min_error_to_stop = 0.01, 
+   init_matrix_value = 1e-6, input_pruning = TRUE, hidden_pruning = TRUE)
R> mod <- mlp(x, y, size = 5, pruneFunc = "OptimalBrainSurgeon",
+  pruneFuncParams = pruneFuncParams)
R> plotnet(mod, rel_rsc = c(3, 8))
R> plotnet(mod, prune_col = 'lightblue', rel_rsc = c(3, 8))
\end{Sinput}
\end{Schunk}
\begin{figure}[!ht]
\subfloat[][]{
\includegraphics[width=0.48\textwidth, page=1]{figs/plotprune.pdf}
\label{fig:plotprune1}
}
\subfloat[][]{
\includegraphics[width=0.48\textwidth, page=2]{figs/plotprune.pdf}
\label{fig:plotprune2}
}
\caption{A pruned neural network from \pkg{RSNNS} \citep{Bergmeir12} using the `Optimal Brain Surgeon' algorithm described in \citet{Zell98}.  The default plotting behavior of \code{plotnet} is to omit pruned connections (\ref{fig:plotprune1}), whereas they can be viewed as dashed lines by including the \code{prune\_col} argument (\ref{fig:plotprune2}).}
\label{fig:plotprune}
\end{figure}

@@

The primary benefit of visualizing a \ac{nid} with \code{plotnet} is the ability to evaluate network architecture and the variation in connections between the layers.  Although useful as a general tool, the \ac{nid} can be difficult to interpret given the amount of weighted connections in most networks.  Alternative methods to quantitatively describe a neural network deconstruct the model weights to determine variable importance, whereas similar information can only be qualitatively inferred from \code{plotnet}.  Two algorithms for evaluating variable importance are available in \pkg{NeuralNetTools}: Garson's algorithm for relative importance \citep{Garson91,Goh95} and Olden's conection weights algorithm \citep{Olden04}.

>>>>

\documentclass[article]{jss}
\usepackage{thumbpdf} 
\graphicspath{{Figures/}}

@@

\author{Marcus W. Beck\\US Environmental Protection Agency}

\title{\pkg{NeuralNetTools}: Visualization and Analysis Tools for Neural Networks}

\Plainauthor{Marcus W. Beck} 
\Plaintitle{NeuralNetTools: Visualization and Analysis Tools for Neural Networks}

\Shorttitle{\pkg{NeuralNetTools}: Visualization and Analysis Tools for Neural Networks}

\Abstract{ Supervised neural networks have been applied as a machine
  learning technique to identify and predict emergent patterns among
  multiple variables.  A common criticism of these methods is the
  inability to characterize relationships among variables from a
  fitted model. Although several techniques have been proposed to
  ``illuminate the black box'', they have not been made available in an
  open-source programming environment.  This article describes the
  \pkg{NeuralNetTools} package that can be used for the interpretation
  of supervised neural network models created in \proglang{R}.
  Functions in the package can be used to visualize a model using a
  neural network interpretation diagram, evaluate variable importance
  by disaggregating the model weights, and perform a sensitivity
  analysis of the response variables to changes in the input
  variables.  Methods are provided for objects from many of the common
  neural network packages in \proglang{R}, including \pkg{caret},
  \pkg{neuralnet}, \pkg{nnet}, and \pkg{RSNNS}.  The article provides
  a brief overview of the theoretical foundation of neural networks, a
  description of the package structure and functions, and an applied
  example to provide a context for model development with
  \pkg{NeuralNetTools}.  Overall, the package provides a toolset for
  neural networks that complements existing quantitative techniques
  for data-intensive exploration.  }

\Keywords{neural networks, \code{plotnet}, sensitivity, variable
  importance, \proglang{R}}

\Plainkeywords{neural networks, plotnet, sensitivity, variable
  importance, R}

%% \Volume{50}
%% \Issue{9}
%% \Month{June}
%% \Year{2012}
%% \Submitdate{2012-06-04}
%% \Acceptdate{2012-06-04}

\Address{
  Marcus W. Beck\\
  US Environmental Protection Agency\\
  National Health and Environmental Effects Research Laboratory\\
  Gulf Ecology Division, 1 Sabine Island Drive\\
  Gulf Breeze, Florida, 32561, USA\\
  Telephone: +1/850/934-2480\\
  E-mail: \email{beck.marcus@epa.gov}
}

@@

@@

A common objective of data-intensive analysis is the synthesis of
unstructured information to identify patterns or trends ``born from the
data'' \citep{Bell09,Kelling09,Michener12}.  Analysis is primarily
focused on data exploration and prediction as compared to
hypothesis-testing using domain-specific methods for scientific
exploration \citep{Kell03}.  Demand for quantitative toolsets to
address challenges in data-rich environments has increased drastically
with the advancement of techniques for rapid acquisition of
data. Fields of research characterized by high-throughput data (e.g.,
bioinformatics; \citealt{Saeys07}) have a strong foundation in
computationally-intensive methods of analysis, whereas disciplines
that have historically been limited by data quantity (e.g., field
ecology; \citealt{Swanson15}) have also realized the importance of
quantitative toolsets given the development of novel techniques to
acquire information.  Quantitative methods that facilitate inductive
reasoning can serve a complementary role to conventional,
hypothesis-driven approaches to scientific discovery \citep{Kell03}.

Statistical methods that have been used to support data exploration
are numerous \citep{Jain00,Recknagel06,Zuur10}.  A common theme among
data intensive methods is the use of machine-learning algorithms where
the primary objective is to identify emergent patterns with minimal
human intervention.  Neural networks, in particular, are designed to
mimic the neuronal structure of the human brain by ``learning'' inherent
data structures through adaptive algorithms
\citep{Rumelhart86,Ripley96}.  Although the conceptual model was
introduced several decades ago \citep{McCulloch43}, neural networks
have had a central role in emerging fields focused on data
exploration.  The most popular form of neural network is the
feed-forward \ac{mlp} trained using the backpropagation algorithm
\citep{Rumelhart86}.  This model is typically used to predict the
response of one or more variables given one to many explanatory
variables.  The hallmark feature of the \ac{mlp} is the
characterization of relationships using an arbitrary number of
parameters (i.e., the hidden layer) that are chosen through iterative
training with the backpropagation algorithm.  Conceptually, the
\ac{mlp} is a hyper-parameterized non-linear model that can fit a
smooth function to any dataset with minimal residual error
\citep{Hornik91}.

An arbitrarily large number of parameters to fit a neural network
provides obvious predictive advantages, but complicates the extraction
of model information.  Diagnostic information such as variable
importance or model sensitivity are necessary aspects of exploratory
data analysis that are not easily obtained from a neural network. As
such, a common criticism is that neural networks are ``black boxes''
that offer minimal insight into relationships among variables
\citep[e.g.,][]{Paruelo97}.  \citet{Olden02} provide a rebuttal to
this concern by describing methods to extract information about
variable relationships from neural networks.  Many of these methods
were previously described but not commonly used.  For example,
\citet{Olden02} describe the \ac{nid} for plotting \citep{Ozesmi99},
the Garson algorithm for variable importance \citep{Garson91}, and the
profile method for sensitivity analysis \citep{Lek96}.  These
quantitative tools ``illuminate the black box'' by disaggregating the
network parameters to characterize relationships between variables
that are described by the model.  Although \ac{mlp} neural networks
were developed for prediction, methods described in \citet{Olden02}
leverage these models to describe data signals.  Increasing the
accessibility of these diagnostic tools will have value for exploratory
data analysis and may also inform causal inference.

This article describes the \pkg{NeuralNetTools} package
\citep{NeuralNetTools} for \proglang{R} \citep{R} that was developed
to better understand information obtained from the \ac{mlp} neural
network.  Functions provided by the package are those described in
\citet{Olden02} but have not been previously available in an
open-source programming environment.  The reach of the package is
extensive in that generic functions were developed for model objects
from the most popular neural network packages available in
\proglang{R}.  The objectives of this article are to 1) provide an
overview of the statistical foundation of the \ac{mlp} network, 2)
describe the theory and application of the main functions in the
\pkg{NeuralNetTools} package, and 3) provide an applied example using
neural networks and \pkg{NeuralNetTools} in data exploration.  The
current released package version is available from the \ac{CRAN}
(CRAN) at \url{https://CRAN.R-project.org/package=NeuralNetTools},
whereas the development version is maintained as a GitHub repository.

\section[Theoretical foundation and existing R packages]{Theoretical foundation and existing \proglang{R} packages}

The typical \ac{mlp} network is composed of multiple layers that
define the transfer of information between input and response layers.
Information travels in one direction where a set of values for
variables in the input layer propagates through one or more hidden
layers to the final layer of the response variables. Hidden layers
between the input and response layers are key components of a neural
network that mediate the transfer of information.  Just as the input
and response layers are composed of variables or nodes, each hidden
layer is composed of nodes with weighted connections that define the
strength of information flow between layers.  Bias layers connected to
hidden and response layers may also be used that are analogous to
intercept terms in a standard regression model.

Training a neural network model requires identifying the optimal
weights that define the connections between the model layers.  The
optimal weights are those that minimize prediction error for a test
dataset that is independent of the training dataset.  Training is
commonly achieved using the backpropagation algorithm described in
\citet{Rumelhart86}.  This algorithm identifies the optimal weighting
scheme through an iterative process where weights are gradually
changed through a forward- and backward-propagation process
\citep{Rumelhart86,Lek00}.  The algorithm begins by assigning an
arbitrary weighting scheme to the connections in the network, followed
by estimating the output in the response variable through the
forward-propagation of information through the network, and finally
calculating the difference between the predicted and actual value of
the response.  The weights are then changed through a backpropagation
step that begins by changing weights in the output layer and then the
remaining hidden layers.  The process is repeated until the chosen
error function is minimized, as in standard model-fitting techniques
for regression \citep{Cheng94}.  A fitted \ac{mlp} neural network can
be represented as \citep{Bishop95,Venables02}:
\begin{equation}
y_k = f_o \left(\sum\limits_{h} w_{hk}f_h \left( \sum\limits_{i} w_{ih}x_i\right) \right),
\end{equation}
where the estimated value of the response variable $y_k$ is a sum of
products between the respective weights $w$ for $i$ input variables
$x$ and $h$ hidden nodes, mediated by the activation functions $f_h$
and $f_o$ for each hidden and output node.

Methods in \pkg{NeuralNetTools} were written for several \proglang{R}
packages that can be used to create \ac{mlp} neural networks:
\pkg{neuralnet} \citep{Fritsch12}, \pkg{nnet} \citep{Venables02}, and
\pkg{RSNNS} \citep{Bergmeir12}. Limited methods were also developed
for neural network objects created with the \code{train} function from
the \pkg{caret} package \citep{Kuhn15}.  Additional \proglang{R}
packages that can create \ac{mlp} neural networks include \pkg{AMORE}
that implements the ``TAO-robust backpropagation algorithm'' for model
fitting \citep{Castejon14}, \pkg{FCNN4R} that provides an \proglang{R}
interface to the \pkg{FCNN} \proglang{C++} library \citep{Klima15},
\pkg{monmlp} for networks with partial monotonicity constraints
\citep{Cannon15}, and \pkg{qrnn} for quantile regression neural
networks \citep{Cannon11}.  At the time of writing, the \ac{CRAN}
download logs \citep{Csardi15} showed that the \proglang{R} packages
with methods in \pkg{NeuralNetTools} included 95\% of all downloads
for the available \ac{mlp} packages, with \pkg{nnet} accounting for
over 78\%.  As such, methods have not been included in
\pkg{NeuralNetTools} for the remaining packages, although further
development of \pkg{NeuralNetTools} could include additional methods
based on popularity.  Methods for each function are currently
available for `\code{mlp}' (\pkg{RSNNS}), `\code{nn}'
(\pkg{neuralnet}), `\code{nnet}' (\pkg{nnet}), and `\code{train}'
(\pkg{caret}; only if the object also inherits from the `\code{nnet}'
class) objects.  Additional default methods or methods for class
`\code{numeric}' are available for some of the generic functions.

@@

The stable release of \pkg{NeuralNetTools} can be installed from \ac{CRAN} and loaded as follows:
%
\begin{Schunk}
\begin{Sinput}
R> install.packages("NeuralNetTools")
R> library("NeuralNetTools")
\end{Sinput}
\end{Schunk}
%
\pkg{NeuralNetTools} includes four main functions that were developed
following similar techniques in \citet{Olden02} and references
therein.  The functions include \code{plotnet} to plot a neural
network interpretation diagram, \code{garson} and \code{olden} to
evaluate variable importance, and \code{lekprofile} for a sensitivity
analysis of neural network response to input variables.  Most of the
functions require the extraction of model weights in a common format
for the neural network object classes in \proglang{R}.  The
\code{neuralweights} function can be used to retrieve model weights
for any of the model classes described above.  A two-element
\code{list} is returned with the first element describing the
structure of the network (number of nodes in the input, hidden, and
output layers) and the second element as a named list of model
weights.  The function is used internally within the main functions
but may also be useful for comparing networks of different classes.

A common approach for data pre-processing is to normalize the input
variables and to standardize the response variables
\citep{Lek00,Olden02}.  A sample dataset that follows this format is
included with \pkg{NeuralNetTools}.  The \code{neuraldat} dataset is a
simple \code{data.frame} with 2000 rows of observations and five
columns for two response variables (\code{Y1} and \code{Y2}) and three
input variables (\code{X1}, \code{X2}, and \code{X3}).  The input
variables are random observations from a standard normal distribution
and the response variables are linear combinations of the input
variables with additional random components.  The response variables
are also scaled from zero to one.  Variables in additional datasets can be
pre-processed to this common format using the \code{scale} function
from \pkg{base} to center and scale input variables (i.e., z-scores) 
and the \code{rescale} function from \pkg{scales} to scale response variables 
from zero to one.  The examples below use three models created from the 
\code{neuraldat} dataset and include `\code{mlp}' (\pkg{RSNNS}), `\code{nn}' 
(\pkg{RSNNS}), and `\code{nnet}' (\pkg{nnet}) objects.
%
\begin{Schunk}
\begin{Sinput}
R> set.seed(123)
R> library("RSNNS")
R> x <- neuraldat[, c("X1", "X2", "X3")]
R> y <- neuraldat[, "Y1"]
R> mod1 <- mlp(x, y, size = 5)
R> library("neuralnet")
R> mod2 <- neuralnet(Y1 ~ X1 + X2 + X3, data = neuraldat, hidden = 5)
R> library("nnet")
R> mod3 <- nnet(Y1 ~ X1 + X2 + X3, data = neuraldat, size = 5)
\end{Sinput}
\end{Schunk}

@@

The \code{plotnet} function in \pkg{NeuralNetTools} plots a \acl{nid}
\citep[\acs{nid};][]{Ozesmi99} and includes several options to
customize aesthetics. A \ac{nid} is a modification of the standard
conceptual illustration of the \ac{mlp} network that changes the
thickness and color of the weight connections based on magnitude and
sign, respectively.  Positive weights between layers are shown as
black lines and negative weights as gray lines. Line thickness is
proportional to the absolute magnitude of each weight
(\cref{fig:plotnet}).

\begin{figure}[t!]
\subfloat[][]{
\includegraphics[width=7.5cm,page=1, trim = 0 20 0 15, clip]{plotnet.pdf}
\label{fig:plotnet1}
}
\subfloat[][]{
\includegraphics[width=7.5cm, page=2, trim = 0 20 0 15, clip]{plotnet.pdf}
\label{fig:plotnet2}
}
\caption{Examples from the \code{plotnet} function showing neural networks as a standard graphic (\ref{fig:plotnet1}) and using the \acl{nid} (\ref{fig:plotnet2}).  Labels outside of the nodes represent variable names and labels within the nodes indicate the layer and node (I: input, H: hidden, O: output, B: bias).}
\label{fig:plotnet}
\end{figure}

A primary and skip layer network can also be plotted for `\code{nnet}'
models with a skip layer connection (\cref{fig:plotnet_skip}). Models
with skip layers include additional connections from the input to
output layers that bypass the hidden layer \citep{Ripley96}.  The
default behavior of \code{plotnet} is to plot the primary network,
whereas the skip layer can be viewed separately with \code{skip =
  TRUE}. If \code{nid = TRUE}, the line widths for both the primary
and skip layer plots are relative to all weights. Plotting a network
with only a skip layer (i.e., no hidden layer, \code{size = 0} in
\code{nnet}) will include bias connections to the output layer,
whereas these are included only in the primary plot if size is greater
than zero.

\begin{figure}[t!]
\subfloat[][]{
\includegraphics[width=7.5cm,page=1, trim = 0 20 0 15, clip]{plotnet_skip.pdf}
\label{fig:plotnet_skip1}
}
\subfloat[][]{
\includegraphics[width=7.5cm, page=2, trim = 0 20 0 15, clip]{plotnet_skip.pdf}
\label{fig:plotnet_skip2}
}
\caption{Examples from the \code{plotnet} function showing a neural network with a separate skip layer between the input and output layers.  The skip layer (\ref{fig:plotnet_skip1}) and primary neural network (\ref{fig:plotnet_skip2}) can be viewed separately with \code{plotnet} by using \code{skip = TRUE} or \code{skip = FALSE}.}
\label{fig:plotnet_skip}
\end{figure}

The \pkg{RSNNS} package provides algorithms to prune connections or
nodes in a neural network \citep{Bergmeir12}.  This approach can
remove connection weights between layers or input nodes that do not
contribute to the predictive performance of the network.  In addition
to visualizing connections in the network that are not important,
connections that are pruned can be removed in successive model
fitting.  This reduces the number of free parameters (weights) that
are estimated by the model optimization algorithm, increasing the
likelihood of convergence to an estimable numeric solution for the
remaining connection weights that minimizes prediction error
\citep[i.e., model identifiability;][]{Ellenius00}. Algorithms in
\pkg{RSNNS} for weight pruning include magnitude-based pruning,
optimal brain damage, and optimal brain surgeon, whereas algorithms
for node pruning include skeletonization and the non-contributing
units method \citep{Zell98}.  The \code{plotnet} function can plot a
pruned neural network, with options to omit or display the pruned
connections (\cref{fig:plotprune}).
%
\begin{Schunk}
\begin{Sinput}
R> pruneFuncParams <- list(max_pr_error_increase = 10.0,
+    pr_accepted_error = 1.0, no_of_pr_retrain_cycles = 1000, 
+    min_error_to_stop = 0.01, init_matrix_value = 1e-6,  
+    input_pruning = TRUE, hidden_pruning = TRUE)
R> mod <- mlp(x, y, size = 5, pruneFunc = "OptimalBrainSurgeon",
+    pruneFuncParams = pruneFuncParams)
R> plotnet(mod, rel_rsc = c(3, 8))
R> plotnet(mod, prune_col = "lightblue", rel_rsc = c(3, 8))
\end{Sinput}
\end{Schunk}
%
\begin{figure}[t!]
\subfloat[][]{
\includegraphics[width=0.48\textwidth, page=1, trim = 0 20 0 65, clip]{plotprune.pdf}
\label{fig:plotprune1}
}
\subfloat[][]{
\includegraphics[width=0.48\textwidth, page=2, trim = 0 20 0 65, clip]{plotprune.pdf}
\label{fig:plotprune2}
}
\caption{A pruned neural network from \pkg{RSNNS} \citep{Bergmeir12}
  using the ``optimal brain surgeon'' algorithm described in
  \citet{Zell98}.  The default plotting behavior of \code{plotnet} is
  to omit pruned connections (\ref{fig:plotprune1}), whereas they can
  be viewed as dashed lines by including the \code{prune\_col}
  argument (\ref{fig:plotprune2}).}
\label{fig:plotprune}
\end{figure}

@@

The primary benefit of visualizing a \ac{nid} with \code{plotnet} is the ability to evaluate network architecture and the variation in connections between the layers.  Although useful as a general tool, the \ac{nid} can be difficult to interpret given the amount of weighted connections in most networks.  Alternative methods to quantitatively describe a neural network deconstruct the model weights to determine variable importance, whereas similar information can only be qualitatively inferred from \code{plotnet}.  Two algorithms for evaluating variable importance are available in \pkg{NeuralNetTools}: Garson's algorithm for relative importance \citep{Garson91,Goh95} and Olden's connection weights algorithm \citep{Olden04}.
